### 2021-05-14 compound_criterion.tex
First draft (s.t. examples)

Standard design optimality theory is developed under the assumption that the fitted model represents the true relationship of interest and that there is no misspecification. However, it is obviously quite a strong belief and in reality we need to take into account at least the possibility that the model might not provide a good fit to the data. In Section \ref{sec::back_misspecification} a brief overview of various types of model misspecification is presented.

In this chapter we consider the case when the fitted polynomial model is nested within a larger model that is assumed to provide a better fit for the data:
\begin{equation}
\label{eq::full_model}
\bm{Y}=\bm{X}_1\bm{\beta}_1+\bm{X}_2\bm{\beta}_2+\bm{\varepsilon},
\end{equation}
where $\bm{X}_2$ is an $n\times q$ is an extension of the initial model matrix representing those extra $q$ potential terms that might represent the fitted model disturbance. The vector $\bm{\beta}_2$ denotes the corresponding parameters. This larger model has $p+q$ parameters in total and not all of them are estimable when the experiment is relatively small: $n<p+q$; this is the case we consider here.
\subsection{Lack-of-fit criterion}
We consider the Generalised criterion presented by Goos et al.: [more details?]
\begin{equation*}
\mbox{minimise }|\bm{X}'_{1}\bm{X}_1|^{-\frac{\kappa_{D}}{p}}\times \left|\bm{L}+\frac{\bm{I}_{q}}{\tau^{2}}\right|^{-\frac{\kappa_{LoF}}{q}} \times |\bm{A}'\bm{A}+\bm{I}_{q}|.^{\frac{\kappa_{bias}}{q}},
\end{equation*}
where $\bm{L}$ is known in literature as dispersion matrix, and the matrix $\bm{L}+\bm{I}_{q}/\tau^{2}$ in the second component is essentially the inverse posterior variance-covariance matrix of the vector of potential terms (up to a factor of $\sigma^2$).

 ** Derivation first or the Generalised criterion

A diffuse prior shall be put on primary terms (as was done by \cite{DuMouchel1994}) -- an arbitrary mean and a variance going to infinity, and the prior on potential terms is the one specified before: $\bm{\beta}_2\sim\mathcal{N}(0,\bm{\Sigma}_{0})$,
$\bm{\Sigma}_{0}=\sigma^{2}\tau^{2}\bm{I}_{q}$, then the posterior
distribution of the coefficients is (as in \cite{Koch2007introduction} and \cite{DuMouchel1994}):
$$\bm{\beta}|\bm{Y}\sim \mathcal{N}(\bm{b},\bm{\Sigma}),\mbox{ where }\bm{b}=\bm{\Sigma X}'\bm{Y}, \bm{X}=[\bm{X}_1, \bm{X}_2],$$
$$\bm{\Sigma}=\left[\frac{\bm{K}}{\sigma^{2}\tau^{2}}+\sigma^{-2}(\bm{X'X})\right]^{-1}=\sigma^{2}[\bm{K}/\tau^{2}+\bm{X'}\bm{X}],^{-1}$$
and 
\begin{equation*}
\bm{K}=\begin{pmatrix}
\bm{0}_{p\times p} & \bm{0}_{p\times q}\\
\bm{0}_{q\times p} & \bm{I}_{q\times q}
\end{pmatrix}.
\end{equation*}
The marginal posterior of $\bm{\beta}_2$ is $\mathcal{N}(\bm{b}_{2},\bm{\Sigma}_{22})$, so using the general formula for the inverse of a block matrix
$$\begin{bmatrix}
 \bm{A}& \bm{B}\\
 \bm{C}& \bm{D}
\end{bmatrix}^{-1}=\begin{bmatrix}
\ldots & \ldots\\
\ldots & (\bm{D}-\bm{CA}^{-1}\bm{B})^{-1}
\end{bmatrix},$$
we can obtain the expression for $\bm{\Sigma}_{22}$:
\begin{align*}
[\bm{K}/\tau^{2}+\bm{X}'\bm{X}]^{-1}&=\begin{bmatrix}
 \bm{X}'_1\bm{X}_1& \bm{X}'_1\bm{X}_2 \\
 \bm{X}'_2\bm{X}_1& \bm{X}'_2\bm{X}_2+\bm{I}_{q}/\tau^{2}
\end{bmatrix}^{-1}\\&=\begin{bmatrix}
\ldots & \ldots\\
\ldots &
(\bm{X}'_2\bm{X}_2+\bm{I}_{q}/\tau^{2}-\bm{X}'_2\bm{X}_1(\bm{X}'_1\bm{X}_1)^{-1}\bm{X}'_1\bm{X}_2)^{-1}
\end{bmatrix},\\
\bm{\Sigma}_{22}&=\sigma^{2}[(\bm{K}/\tau^{2}+\bm{X}^{'}\bm{X})^{-1}]_{22}=\sigma^{2}\left(\bm{L}+\frac{\bm{I}_{q}}{\tau^{2}}\right)^{-1},
\end{align*} 
and $$(\bm{\Sigma}_{22})^{-1}=\frac{1}{\sigma^{2}}\left(\bm{L}+\frac{\bm{I}_{q}}{\tau^{2}}\right).$$

Therefore, the $(1-\alpha)\times100\%$ confidence region for the potential terms over the posterior distribution, when $\sigma^2$ is estimated by $s^2$ on $\nu$ degrees of freedom, is given by \citep{Draper1998}:

$$(\bm{\beta}_{2}-\bm{b}_{2})^{'}(\bm{L}+\bm{I}_{q}/\tau^{2})(\bm{\beta}_{2}-\bm{b}_{2})\leq qs^{2}F_{q,\nu;1-\alpha},$$
where $\mathrm{F}(q,\nu,\alpha)$ is the $\alpha$-quantile of F-distribution with $q$ and $\nu$ degrees of freedom.\\

\subsection{MSE-based bias criterion}
From the inference point of view, in the case of potential model contamination the bias in the parameters' estimates would be of substantial interest. A natural way of evaluating the quality of $\bm{\hat{\beta}}_{1}$, as noted by \citet{FedorovMontepiedra1997} is the matrix of  mean square error, which is the $\bm{L}_2$-distance between the true and estimated values with respect to the probability distribution measure of $\bm{Y}$ under the assumption of model (\ref{eq::full_m}):
\begin{align}
\label{eq::MSE}
\mbox{MSE}(\bm{\hat{\beta}}_1|\bm{\beta})=&\mathtt{E}_{\bm{Y}|\bm{\beta}}[(\bm{\hat{\beta}}_1-\bm{\beta}_1)(\bm{\hat{\beta}}_1-\bm{\beta}_1)']\notag\\=&\sigma^2(\bm{X}_1^{'}\bm{X}_1)^{-1}+\bm{A}\bm{\beta}_2\bm{\beta}_2'\bm{A}', 
\end{align}
where, as before, $\bm{A}=(\bm{X}_1^{'}\bm{X}_1)^{-1}\bm{X}_1^{'}\bm{X}_2$ is the alias matrix.
The determinant-based criterion function is minimising
\begin{equation}
\label{eq::MSE_det}
\mathtt{E}_{\beta}\log(\det[\mbox{MSE}(\bm{\hat{\beta}}_1|\bm{\beta})]),
\end{equation}
where by $\bm{\beta}$ we denote the joint vector of primary and potential model parameters: $\bm{\beta}=[\bm{\beta}_1, \bm{\beta}_2]$.
Using the matrix determinant lemma \citep{Harville2006matrix}
\begin{equation*}
\det[\bm{P}+\bm{uv}']=\det[\bm{P}]\det[1+\bm{v'}\bm{P}^{-1}\bm{u}], 
\end{equation*}
where $\bm{P}$ is an invertible square matrix and $\bm{u}$, $\bm{v}$ are column vectors, the determinant in (\ref{eq::MSE_det}) can be decomposed:
\begin{align}
\label{eq::mse_det_dec}
&\det[\mbox{MSE}(\bm{\hat{\beta}}_1|\bm{\beta}_2)]\notag\\&=\det[\sigma^2\bm{M}^{-1}+\bm{M}^{-1}\bm{X}_1^{'}\bm{X}_2\bm{\beta}_2\bm{\beta}_2'\bm{X}_2^{'}\bm{X}_1\bm{M}^{-1}]\notag\\&=\sigma^{2p}\det[\bm{M}^{-1}+\bm{M}^{-1}\bm{X}_1^{'}\bm{X}_2\bm{\tilde{\beta}}_2\bm{\tilde{\beta}}_2'\bm{X}_2^{'}\bm{X}_1\bm{M}^{-1}]\notag\\&=\sigma^{2p}\det[\bm{M}^{-1}]\det[1+\bm{\tilde{\beta}}_2'\bm{X}_2^{'}\bm{X}_1\bm{M}^{-1}\bm{M}\bm{M}^{-1}\bm{X}_1^{'}\bm{X}_2\bm{\tilde{\beta}}_2]\notag\\&=\sigma^{2p}\det[\bm{M}^{-1}](1+\bm{\tilde{\beta}}_2'\bm{X}_2^{'}\bm{X}_1\bm{M}^{-1}\bm{X}_1^{'}\bm{X}_2\bm{\tilde{\beta}}_2),\\
&\mbox{where } \bm{M}=\bm{X}_1^{'}\bm{X}_1 \mbox{ and }\bm{\tilde{\beta}}_2=\bm{\beta}_2/\sigma. \notag
\end{align}
Applying the logarithm function gives us the following sum:
\begin{align*}
&\log(\det[\mbox{MSE}(\bm{\hat{\beta}}_1|\bm{\beta}_2)])\\&=p\log\sigma^2+\log(\det[\bm{M}^{-1}])+\log(1+\bm{\tilde{\beta}}_2'\bm{X}_2^{'}\bm{X}_1\bm{M}^{-1}\bm{X}_1^{'}\bm{X}_2\bm{\tilde{\beta}}_2). 
\end{align*}
The first summand does not depend on the design, so it will not be included in the criterion; the second one is the $D$-optimality component. Therefore, (\ref{eq::MSE_det}) is replaced by
\begin{equation*}
\log(\det[\bm{M}^{-1}])+\mathtt{E}_{\bm{\tilde{\beta}}_2}\log(1+\bm{\tilde{\beta}}_2'\bm{X}_2^{'}\bm{X}_1\bm{M}^{-1}\bm{X}_1^{'}\bm{X}_2\bm{\tilde{\beta}}_2).
\end{equation*}

The second term needs to be evaluated numerically due to the obvious lack of information regarding the potential terms' coefficients. Note that as $\bm{\beta}_2 \sim \mathcal{N}(\bm{0},\tau^{2}\sigma^{2}\bm{I}_{q})$, then $\bm{\tilde{\beta}}_2 \sim \mathcal{N}(\bm{0},\tau^{2}\bm{I}_{q})$ and its prior distribution does not depend on the unknown $\sigma^2$.

To sum up, the resulting criterion function comprising the $DP$-criterion and the $LoF(DP)$-component from the generalised $DP$-criterion as well as the MSE-based part, all brought to the scale of efficiencies, is
\begin{align}
\label{eq::MSE_D}
\mbox{minimise }&\left[\left|\bm{X}'_{1}\bm{X}_{1}\right|^{-1/p}F_{p,d;1-\alpha_{DP}}\right]^{\kappa_{DP}} \times \notag \\ &\left[\left|\bm{L}+\frac{\bm{I}_{q}}{\tau^{2}}\right|^{-1/q}F_{q,d;1-\alpha_{LoF}}\right]^{\kappa_{LoF}}\times \notag\\ & \left[|\bm{X}'_{1}\bm{X}_{1}|^{-1}\exp\left(\frac{1}{N}\sum_{i=1}^{N}\log(1+\bm{\tilde{\beta}}_{2i}'\bm{X}_2^{'}\bm{X}_1\bm{M}^{-1}\bm{X}_1^{'}\bm{X}_2\bm{\tilde{\beta}}_{2i})\right)\right]_{.}^{\kappa_{MSE}/p}
\end{align}
Here $\bm{\tilde{\beta}}_{2i}$, $i=1..N$ -- independent random vectors sampled from $\mathcal{N}(\bm{0},\tau^{2}\sigma^{2}\bm{I}_{q})$.

\subsubsection{Point-based estimate}

One of the alternatives to the MC approach we would be willing to consider is to use the point prior for $\bm{\beta}_2$, that is $\bm{\beta}_2=\sigma\tau\bm{1}_q$ (where $\bm{1}_q$ is a $q$-dimensional vector with all components being equal to $1$), which is the standard deviation  of the initial (normal) prior. Therefore, $\bm{\tilde{\beta}}_2=\tau\bm{1}_q$, and the MSE(D)-part is
\begin{align*}
&|\bm{X}'_{1}\bm{X}_{1}|^{-1}(1+\bm{\tilde{\beta}}_{2}'\bm{X}_2^{'}\bm{X}_1\bm{M}^{-1}\bm{X}_1^{'}\bm{X}_2\bm{\tilde{\beta}}_{2})\\=&|\bm{X}'_{1}\bm{X}_{1}|^{-1}(1+\tau^2\bm{1}'_q\bm{X}_2^{'}\bm{X}_1\bm{M}^{-1}\bm{X}_1^{'}\bm{X}_2\bm{1}_q)\\=&|\bm{X}'_{1}\bm{X}_{1}|^{-1}\left(1+\tau^2\sum_{i,j=1}^{q}[\bm{X}_2^{'}\bm{X}_1\bm{M}^{-1}\bm{X}_1^{'}\bm{X}_2]_{(i,j)}\right),
\end{align*}
so essentially the last part of the equation is the summation of matrix elements, which obviously takes less computational time even for large dimensions $q$.

The normal prior for $\bm{\beta}_2$ shall be preserved in the lack-of-fit component in (\ref{eq::MSE_D}), and so the criterion with the point prior estimate of the MSE(D)-component is the following function:
\begin{align}
\label{eq::MSE_D_point}
\mbox{minimise }&\left[\left|\bm{X}'_{1}\bm{X}_{1}\right|^{-1/p}F_{p,d;1-\alpha_{DP}}\right]^{\kappa_{DP}} \times \notag \\ &\left[\left|\bm{L}+\frac{\bm{I}_{q}}{\tau^{2}}\right|^{-1/q}F_{q,d;1-\alpha_{LoF}}\right]^{\kappa_{LoF}}\times \notag\\ & \left[|\bm{X}'_{1}\bm{X}_{1}|^{-1}\left(1+\tau^2\sum_{i,j=1}^{q}[\bm{X}_2^{'}\bm{X}_1\bm{M}^{-1}\bm{X}_1^{'}\bm{X}_2]_{(i,j)}\right)\right]_{.}^{\kappa_{MSE}/p}
\end{align}

Later we will refer to this as the ``MSE(D)P''-criterion, and in Section \ref{sec::mse_example} we will explore how much efficiency is lost when it is used instead of the computationally expensive original MSE(D)-based criterion (\ref{eq::MSE_D}).

To infer the trace-based form of the criterion, we use the pseudo-Bayesian approach to calculating the trace function of the MSE matrix (\ref{eq::MSE}): 
\begin{align*}
\mathtt{E}_{\beta_2}\mbox{trace}[\mbox{MSE}(\bm{\hat{\beta}}_1|\bm{\beta}_2)]&=\mbox{trace}[\mathtt{E}_{\beta_2}\mbox{MSE}(\bm{\hat{\beta}}_1|\bm{\beta}_2)]=\\&=\mbox{trace}[\sigma^2(\bm{X}_1^{'}\bm{X}_1)^{-1} + \mathtt{E}_{\bm{\beta}_2}(\bm{A}\bm{\beta}_2\bm{\beta}_2'\bm{A}')]=\\&=\mbox{trace}[\sigma^2(\bm{X}_1^{'}\bm{X}_1)^{-1}+\sigma^2\tau^2\bm{A}\bm{A}']=\\&=\sigma^2\mbox{trace}[(\bm{X}_1^{'}\bm{X}_1)^{-1}+\tau^2\bm{A}\bm{A}']\\&=\sigma^2[\mbox{trace}\{(\bm{X}_1^{'}\bm{X}_1)^{-1}\}+\tau^2\mbox{trace}\bm{A}\bm{A}'].
\end{align*}

The transitions above are possible due to commutativity of the operations of trace and expectation. The main advantage is the absence of the necessity of any additional numerical evaluations, and in this case of the trace-based criterion using the point prior for $\bm{\beta}_2$ defined earlier would lead to the same function. By minimising the whole function the sum of primary terms' variance and the expected squared norm of the bias vector in the direction of the potential terms are minimised simultaneously. The second part here contains the scaling parameter $\tau^2$ which regulates the magnitude of the potential terms' variation relatively to the error variance.

Joining up the derived criterion function, the $LP$-criterion and the $LoF(LP)$-component from the generalised $LP$-criterion (\ref{eq::GLP_eff}), we obtain the compound criterion:
\begin{align}
\label{eq::MSE_L}
\mbox{minimise} &\left[\frac{1}{p}\mbox{trace}(\bm{WX}'_{1}\bm{X}_{1})^{-1}F_{1,d;1-\alpha_{LP}}\right]^{\kappa_{LP}}\times \notag\\& \left[\frac{1}{q}\mbox{trace}\left(\bm{L}+\frac{\bm{I}_{q}}{\tau^{2}}\right)^{-1}F_{1,d;1-\alpha_{LoF}}\right]^{\kappa_{LoF}}\times 
\notag\\& \left[\frac{1}{p}\mbox{trace}\{(\bm{X}'_{1}\bm{X}_{1})^{-1}+\tau^2\bm{A}\bm{A}'\}\right]^{\kappa_{MSE}}_{.}
\end{align}

>>>>

Standard design optimality theory is developed under the assumption that the primary model provides the most proper fit for the data: in many real applications this is quite a strong belief, and in reality we need to take into account at least the possibility that some misspecification is present at the planning stage.
%In Section \ref{sec::back_misspecification} a brief overview of various types of model misspecification is presented. 
%Other types of model misspecification (in background)

In this work we consider the case when the fitted polynomial model with $p$ parameters is nested within a larger model that is assumed to provide a better fit for the data:
\begin{equation}
\label{eq::full_model}
\bm{Y}=\bm{X}_p\bm{\beta}_p+\bm{X}_q\bm{\beta}_q+\bm{\varepsilon},
\end{equation}
where $\bm{X}_q$ is an $n\times q$ extension of the primary model matrix containing extra $q$  terms that we refer to as "potential terms" and that represent the fitted model disturbance, with vector $\bm{\beta}_q$ denoting the corresponding parameters. They are not of any inferential interest, and, moreover, not all of them are necessarily estimable when the experiment is relatively small, i.e. $n<p+q$ -- this is the case we mainly consider here. As usually, we do assume independent and normally distributed error terms: $\bm{\varepsilon}\sim \mathcal{N}(\bm{0},\sigma^{2}\bm{I}_{n})$.

\subsection*{Lack-of-fit criterion}

Aiming towards controlling the magnitude and scope of the potential terms, we adapt Bayesian approach regarding the full model parameters.  Diffuse prior shall be put on primary terms -- an arbitrary mean and a variance going to infinity, and the prior on potential terms is a normal distribution: $\bm{\beta}_q\sim\mathcal{N}(0,\bm{\Sigma}_{0})$, with the prior variance scaled with respect to the error variation:
$\bm{\Sigma}_{0}=\sigma^{2}\tau^{2}\bm{I}_{q}$. Then, following the normality in the model (\ref{eq::full_model}), the posterior distribution of the joint vector of  coefficients $\bm{\beta}  = [\bm{\beta}^T_p, \bm{\beta}^T_q]^T$ is (\cite{Koch2007introduction} is multivariate normal \cite{DuMouchel1994}):
\begin{align*}
\bm{\beta}|\bm{Y} &\sim \mathcal{N}(\bm{b},\bm{\Sigma}), \\
\mbox{where } \bm{b} = \bm{\Sigma X}^T\bm{Y} &\mbox{ and }  \bm{\Sigma} = \sigma^{2}[\bm{X}^T\bm{X} + \bm{K}/\tau^{2}],^{-1}  \bm{X}=[\bm{X}_p, \bm{X}_q],\\
\bm{K} &= \begin{pmatrix}
\bm{0}_{p\times p} & \bm{0}_{p\times q}\\
\bm{0}_{q\times p} & \bm{I}_{q\times q}
\end{pmatrix}.
\end{align*}

The marginal posterior distribution of $\bm{\beta}_q$ is also multivariate normal with mean $\bm{b}_q$ -- the last $q$ elements of $\bm{b}$ and the covariance matrix $\bm{\Sigma}_{qq}$-- the bottom right $q \times q $ submatrix of $\bm{\Sigma}$:
\begin{align*}
\bm{\Sigma}_{qq} = \sigma^{2}[(\bm{X}^T\bm{X} +  \bm{K}/\tau^{2})^{-1}]_{[q,q]} &= 
\sigma^{2}\begin{bmatrix}
\bm{X}^T_p\bm{X}_p& \bm{X}^T_p\bm{X}_q \\
\bm{X}^T_q\bm{X}_p& \bm{X}^T_q\bm{X}_q+\bm{I}_{q}/\tau^{2}
\end{bmatrix}^{-1}_{[q,q]}\\&=
\sigma^{2}[\bm{X}^T_q\bm{X}_q+\bm{I}_{q}/\tau^{2}-\bm{X}^T_q\bm{X}_p(\bm{X}^T_p\bm{X}_p)^{-1}\bm{X}^T_p\bm{X}_q]^{-1}\\&=\sigma^{2}\left[\bm{L}+\bm{I}_{q}/\tau^{2}\right],^{-1}
\end{align*} 
where $\bm{L} = \bm{X}^T_q\bm{X}_q-\bm{X}^T_q\bm{X}_p(\bm{X}^T_p\bm{X}_p)^{-1}\bm{X}^T_p\bm{X}_q$ is known in model-sensitivity design literature as the ``dispersion matrix'' [references], which provides a point-wise measure of the distance between column vectors of $\bm{X}_q$ and linear subspace defined by the column vectors of the primary model matrix $\bm{X}_p$, that is how well each of the potential terms could be approximated by a linear span of the primary ones [?].

Reducing the primary model's lack-of-fit in the direction of the potential terms can be translated into a criterion function of the design by utilising the posterior distribution for $\bm{\beta}_q$ derived above and constructing a $(1-\alpha)\times100\%$ confidence region for the parameters depending on the model matrices and the variance estimate $s^2$ on $\nu$ degrees of freedom \citep{Draper1998}:
\begin{equation*}
(\bm{\beta}_{2}-\bm{b}_{2})^{'}(\bm{L}+\bm{I}_{q}/\tau^{2})(\bm{\beta}_{2}-\bm{b}_{2})\leq qs^{2}F_{q,\nu;1-\alpha},
\end{equation*}
where $\mathrm{F}_{q,\nu; \alpha}$ is the $\alpha$-quantile of F-distribution with $q$ and $\nu$ degrees of freedom.
Minimising the volume of the confidence region is equivalent to 
\begin{equation}
\label{eq::LoFDP_criterion}
\left|\bm{L}+\bm{I}_{q}/\tau^{2}\right|^{-1/q}F_{q,d;1-\alpha_{LoF}} \longrightarrow \mbox{ min, }  
\end{equation}  
which we refer to as ``Lack-of-fit DP-criterion'', that is directly related to [1] the lack-of-fit component in the Generalised D-optimality developed by \cite{Goos2005model} -- where the residual number of degrees of freedom $\nu$ does not depend on the design, and [2] DP-optimality \citep{GilmourTrinca2012}, with the F-quantile preserved from $\nu = d$ being the number of replicates in the design.

Another way of formulating a criterion that we consider in this work -- minimising the average of posterior variances of linear functions of $\bm{\beta}_q$ defined by matrix $\bm{J}$, and we define the Lack-of-fit LP-criterion as mean of the squared lengths of the $(1-\alpha)\times100\%$ confidence intervals for these linear functions:
\begin{equation}
\label{eq::LoFLP_criterion}
\frac{1}{q}\mbox{trace}\left[\bm{JJ}^T\left(\bm{L}+\frac{\bm{I}_{q}}{\tau^{2}}\right)^{-1}\right]F_{1,d;1-\alpha}  \longrightarrow \mbox{ min. } 
\end{equation}

This trace-based criterion is linked to the lack-of-fit part of the Generalised L-optimality \citep{Goos2005model}, and the pure error estimation approach retains the corresponding F-quantile. 
[Henceforth we mainly consider the case when $\bm{J}$ is the identity matrix, that is we work with the analogue of $AP$-optimality. In other words, the lack-of-fit component in the generalised $AP$-criterion  stands for the minimisation of the $\bm{L}_2$-norm of the $q$-dimensional vector of the posterior confidence intervals' lengths for the potential parameters. ]

\subsection*{MSE-based bias criterion}
Together with controlling the magnitude of model contamination, it is also desirable to "protect" the quality of inference that is to be drawn through fitting the primary model, from the potential presence of extra terms. 
% minimise its effect on the primary model inference, that is the quality of paramters' estimates .
From this point of view, the bias of the parameters' estimates $\hat{\bm{beta}}_p$ would be of substantial interest; a natural way of evaluating their quality is the matrix of  mean square error \citep{FedorovMontepiedra1997} , which is the $\bm{L}_2$-distance between the true and estimated values with respect to the probability distribution measure of $\bm{Y}$ under the assumption of model (\ref{eq::full_model}):
\begin{align}
\label{eq::MSE}
\mbox{MSE}(\bm{\hat{\beta}}_p|\bm{\beta})=&\mathtt{E}_{\bm{Y}|\bm{\beta}}[(\bm{\hat{\beta}}_p-\bm{\beta}_p)(\bm{\hat{\beta}}_p-\bm{\beta}_p)^T]\notag\\=&\sigma^2(\bm{X}_p^T\bm{X}_p)^{-1}+\bm{A}\bm{\beta}_q\bm{\beta}_q^T\bm{A}^T, 
\end{align}
where $\bm{A}=(\bm{X}_p^{T}\bm{X}_p)^{-1}\bm{X}_p^{T}\bm{X}_q$ denotes the $p \times q$ alias matrix, which elements reflect the linearity scale of the relationship between the primary (rows) and potential (columns) terms.

The determinant-based criterion that would correspond to the overall simultaneous minimisation of the bias above is constructed as log-determinant of the MSE matrix averaged across [?] the values of the full model parameters $\bm{\beta}$:
\begin{equation}
\label{eq::MSE_det}
\mathtt{E}_{\bm{\beta}}\log(\det[\mbox{MSE}(\bm{\hat{\beta}}_p|\bm{\beta})]) \longrightarrow \mbox{ min.}
\end{equation}

Using the matrix determinant lemma \citep{Harville2006matrix} and setting $\bm{M}=\bm{X}_1^{'}\bm{X}_1$ and $\bm{\tilde{\beta}}_2=\bm{\beta}_2/\sigma,$ the determinant and the applied logarithm in (\ref{eq::MSE_det})) can be decomposed:
%\begin{equation*}
%\det[\bm{P}+\bm{uv}']=\det[\bm{P}]\det[1+\bm{v'}\bm{P}^{-1}\bm{u}], 
%\end{equation*}
%where $\bm{P}$ is an invertible square matrix and $\bm{u}$, $\bm{v}$ are column vectors, 
\begin{align}
\label{eq::mse_det_dec}
&\det[\mbox{MSE}(\bm{\hat{\beta}}_p|\bm{\beta}_q)]\notag\\&=\det[\sigma^2\bm{M}^{-1}+\bm{M}^{-1}\bm{X}_p^{T}\bm{X}_q\bm{\beta}_q\bm{\beta}_q^T\bm{X}_q^{T}\bm{X}_p\bm{M}^{-1}]\notag\\&=\sigma^{2p}\det[\bm{M}^{-1}+\bm{M}^{-1}\bm{X}_p^{'}\bm{X}_q\bm{\tilde{\beta}}_q\bm{\tilde{\beta}}_q^T\bm{X}_q^{T}\bm{X}_p\bm{M}^{-1}]\notag\\&=\sigma^{2p}\det[\bm{M}^{-1}]\det[1+\bm{\tilde{\beta}}_q^T\bm{X}_q^{T}\bm{X}_p\bm{M}^{-1}\bm{M}\bm{M}^{-1}\bm{X}_p^{T}\bm{X}_q\bm{\tilde{\beta}}_q]\notag\\&=\sigma^{2p}\det[\bm{M}^{-1}](1+\bm{\tilde{\beta}}_q^T\bm{X}_q^{T}\bm{X}_p\bm{M}^{-1}\bm{X}_p^{T}\bm{X}_q\bm{\tilde{\beta}}_q);\\[10pt]
&\log(\det[\mbox{MSE}(\bm{\hat{\beta}}_p|\bm{\beta}_q)])\notag\\&=p\log\sigma^2+\log(\det[\bm{M}^{-1}])+\log(1+\bm{\tilde{\beta}}_q^T\bm{X}_q^{T}\bm{X}_p\bm{M}^{-1}\bm{X}_p^{T}\bm{X}_q\bm{\tilde{\beta}}_q). 
\end{align}
The first summand does not depend on the design, so it will not be included in the criterion; the second one is the $D$-optimality criterion function. Therefore, the criterion in (\ref{eq::MSE_det}) which we refer to becomes
\begin{equation*}
\log(\det[\bm{M}^{-1}])+\mathtt{E}_{\bm{\tilde{\beta}}_q}\log(1+\bm{\tilde{\beta}}_q^T\bm{X}_q^{T}\bm{X}_p\bm{M}^{-1}\bm{X}_p^{T}\bm{X}_q\bm{\tilde{\beta}}_q) \longrightarrow min.
\end{equation*}
Due to the obvious lack of information regarding $\bm{\tilde{\beta}}_q$, the second term needs to be evaluated numerically. Expressing the prior variance of $\bm{\beta}_q$ as a scaled error variance $\bm{\beta}_q \sim \mathcal{N}(\bm{0},\tau^{2}\sigma^{2}\bm{I}_{q})$ means that $\bm{\tilde{\beta}}_q \sim \mathcal{N}(\bm{0},\tau^{2}\bm{I}_{q})$, and that, quite conveniently, its prior distribution does not depend on the unknown $\sigma.^2$ Then a primitive Monte-Carlo can be used to evaluate that term: drawing a sample of large size $N$ from the prior, and approximating the expectation above by the average across the sampled values of $\bm{\tilde{\beta}}_{q_i},$ $i=1,\dots, N$:
\begin{equation*}
\mathtt{E}_{\bm{\tilde{\beta}}_q}\log(1+\bm{\tilde{\beta}}_q^T\bm{X}_q^{T}\bm{X}_p\bm{M}^{-1}\bm{X}_p^{T}\bm{X}_q\bm{\tilde{\beta}}_q) \approx \frac{1}{N}\sum_{i=1}^{N}\log(1+\bm{\tilde{\beta}}_{q_i}^T\bm{X}_q^{T}\bm{X}_p\bm{M}^{-1}\bm{X}_p^{T}\bm{X}_q\bm{\tilde{\beta}}_{q_i}).
\end{equation*}

One of the alternatives to this more computationally demanding approach is to use the point prior for $\bm{\beta}_q$, that is setting  $\bm{\beta}_q=\sigma\tau\bm{1}_q$ (where $\bm{1}_q$ is a $q$-dimensional vector of $1$s), which is the standard deviation  of the initial normal prior. Then $\bm{\tilde{\beta}}_q = \tau\bm{1}_q$ with probability $1$ and the expectation above becomes:
\begin{align*}
\mathtt{E}_{\bm{\tilde{\beta}}_q}\log(1+\bm{\tilde{\beta}}_q^T\bm{X}_q^{T}\bm{X}_p\bm{M}^{-1}\bm{X}_p^{T}\bm{X}_q\bm{\tilde{\beta}}_q)  &\approx \log(1+\tau^2\bm{1}^T_q\bm{X}_q^{T}\bm{X}_p\bm{M}^{-1}\bm{X}_p^{T}\bm{X}_q\bm{1}_q) \\&=\log(1+\tau^2\sum_{i,j=1}^{q}[\bm{X}_q^{T}\bm{X}_p\bm{M}^{-1}\bm{X}_p^{T}\bm{X}_q][i,j]),
\end{align*}
with summation of matrix elements taking less computational time even for large $q$.

To infer the trace-based form of the criterion that would minimise the mean squared bias, we use the pseudo-Bayesian approach of setting priors on the model's parameters  to calculating the trace function of the MSE matrix (\ref{eq::MSE}): 
\begin{align*}
\mathtt{E}_{\bm{\beta}_q}\mbox{trace}[\mbox{MSE}(\bm{\hat{\beta}}_p\bm{\beta}_q)]&=\mbox{trace}[\mathtt{E}_{\bm{\beta}_q}\mbox{MSE}(\bm{\hat{\beta}}_p|\bm{\beta}_q)]=\\&=\mbox{trace}[\sigma^2(\bm{X}_p^{T}\bm{X}_p)^{-1} + \mathtt{E}_{\bm{\beta}_q}(\bm{A}\bm{\beta}_q\bm{\beta}_q^T\bm{A}^T)]=\\&=\mbox{trace}[\sigma^2(\bm{X}_p^{T}\bm{X}_p)^{-1}+\sigma^2\tau^2\bm{A}\bm{A}^T]=\\&=\sigma^2\mbox{trace}[(\bm{X}_p^{T}\bm{X}_p)^{-1}+\tau^2\bm{A}\bm{A}^T]\\&=\sigma^2[\mbox{trace}\{(\bm{X}_p^{T}\bm{X}_p)^{-1}\}+\tau^2\mbox{trace}\bm{A}\bm{A}^T].
\end{align*}

The operations of calculating trace and expectation are commutative, hence there is no necessity of any additional numerical evaluations, and in this case of the trace-based criterion using the point prior for $\bm{\beta}_q$ defined earlier would lead to the same resulting function. By minimising the whole function above, we simultaneously minimise both the sum of primary terms' variance and the expected squared norm of the bias vector in the direction of the potential terms. The second part here contains the scaling parameter $\tau^2$ which regulates the magnitude of the potential terms' variation relatively to the error variance.

\subsection*{Compound criteria}
The compound criterion is constructed to account for the three main objectives -- precision (or accuracy?) of the primary model parameters, control for the lack-of-fit and minimising the inferential bias from the potential model contamination -- as the weighted product of efficiencies (\ref{eq::compound}) with respect to DP-criterion, and Lack-of-Fit DP functions(\ref{eq::LoFDP_criterion}, \ref{eq::LoFLP_criterion}) and the MSE-based ones. All of them are brought to the scale of efficiencies [?and $\exp$]:

\begin{align}
\label{eq::MSE_D}
\mbox{minimise }&\left[\left|\bm{X}^T_{p}\bm{X}_{p}\right|^{-1/p}F_{p,d;1-\alpha_{DP}}\right]^{\kappa_{DP}} \times \notag \\ &\left[\left|\bm{L}+\frac{\bm{I}_{q}}{\tau^{2}}\right|^{-1/q}F_{q,d;1-\alpha_{LoF}}\right]^{\kappa_{LoF}}\times \notag\\ & \left[|\bm{X}^T_{p}\bm{X}_{p}|^{-1}\exp\left(\mathtt{E}_{\bm{\tilde{\beta}}_q}\log(1+\bm{\tilde{\beta}}_q^T\bm{X}_q^{T}\bm{X}_p\bm{M}^{-1}\bm{X}_p^{T}\bm{X}_q\bm{\tilde{\beta}}_q) \right)\right]_{.}^{\kappa_{MSE}/p}
\end{align}

This criterion is later referred to as compound ``MSE-DP-criterion'', where $\alpha_{DP}$ and $\alpha_{LoF}$ denote the confidence intervals' probability levels for the primary $\bm{\beta}_p$ and potential coefficients $\bm{\beta}_q$ -- in the DP and LoF-DP elementary criteria; usually they are set to $0.05$ or $0.01$. As mentioned before, non-negative weights $\kappa_X$ that sum up to $1$ define the compound criterion and are chosen to reflect the experimenter's priorities.
